---
title: "DGE - Unpaired"
author: "Musabbir Sammak"
date: "12/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Loads necessary libraries
```{r, include=FALSE}
library(edgeR)
library(limma)
library(EnsDb.Hsapiens.v86)
library(dplyr)
library(stringr)
library(ggplot2)
library(gplots)
library(ggpubr)
library(RColorBrewer)
library(EnhancedVolcano)
```
```{r}
if (!dir.exists('./Figures')) { dir.create('Figures') }
if (!dir.exists('./Results')) { dir.create('Results') }
```
# Loads & Preprocesses clinical data
```{r}
phenoNormalIlc <- read.csv("./Data/clinical-normal-lobular.csv")
phenoTumorIlc <- read.csv("./Data/clinical-tumor-lobular.csv")
phenoNormalIdc <- read.csv("./Data/clinical-normal-ductal.csv")
phenoTumorIdc <- read.csv("./Data/clinical-tumor-ductal.csv")

phenoNormalIdc$barcode <- str_replace_all(phenoNormalIdc$barcode, "-", ".")
phenoTumorIdc$barcode <- str_replace_all(phenoTumorIdc$barcode, "-", ".")
phenoNormalIlc$barcode <- str_replace_all(phenoNormalIlc$barcode, "-", ".")
phenoTumorIlc$barcode <- str_replace_all(phenoTumorIlc$barcode, "-", ".")
```
```{r}
dim(phenoNormalIlc)
dim(phenoNormalIdc)
dim(phenoTumorIlc)
dim(phenoTumorIdc)
```
# Selects only the unpaired samples
```{r}
pairedPatientsIlc <- intersect(phenoNormalIlc$patient, phenoNormalIlc$patient)
pairedPatientsIdc <- intersect(phenoNormalIdc$patient, phenoNormalIdc$patient)
```
```{r}
phenoTumorIlc <- subset(phenoTumorIlc,
                        !(phenoTumorIlc$patient %in% intersect(pairedPatientsIlc, 
                                                               phenoTumorIlc$patient)))
phenoTumorIdc <- subset(phenoTumorIdc,
                        !(phenoTumorIdc$patient %in% intersect(pairedPatientsIdc,
                                                               phenoTumorIdc$patient)))
```
# Combines tumor and normal clinical data
```{r}
phenoIlc <- rbind(phenoNormalIlc, phenoTumorIlc)
rownames(phenoIlc) <- phenoIlc$barcode
phenoIlc <- phenoIlc[, c("patient", "shortLetterCode")]
colnames(phenoIlc) <- c("Patient", "Group")
phenoIlc$Patient <- factor(phenoIlc$Patient)
phenoIlc$Group <- factor(phenoIlc$Group)
```
```{r}
phenoIdc <- rbind(phenoNormalIdc, phenoTumorIdc)
rownames(phenoIdc) <- phenoIdc$barcode
phenoIdc <- phenoIdc[, c("patient", "shortLetterCode")]
colnames(phenoIdc) <- c("Patient", "Group")
phenoIdc$Patient <- factor(phenoIdc$Patient)
phenoIdc$Group <- factor(phenoIdc$Group)
```
# Preprocess expression data
```{r}
expNormalIlc <- read.csv("./Data/expression-normal-lobular.csv")
expTumorIlc <- read.csv("./Data/expression-tumor-lobular.csv")
expTumorIlc <- expTumorIlc[, intersect(c("X", rownames(phenoIlc)), colnames(expTumorIlc))]

expIlc <- merge(expNormalIlc, expTumorIlc, by = "X")
rownames(expIlc) <- expIlc$X
expIlc <- expIlc[, c(2:length(colnames(expIlc)))]
```
```{r}
expNormalIdc <- read.csv("./Data/expression-normal-ductal.csv")
expTumorIdc <- read.csv("./Data/expression-tumor-ductal.csv")
expTumorIdc <- expTumorIdc[, intersect(c("X", rownames(phenoIdc)), colnames(expTumorIdc))]

expIdc <- merge(expNormalIdc, expTumorIdc, by = "X")
rownames(expIdc) <- expIdc$X
expIdc <- expIdc[, c(2:length(colnames(expIdc)))]
```
# Creates DGEList
```{r}
dgeIlc <- DGEList(counts = as.matrix(expIlc), samples = phenoIlc, group = phenoIlc$Group)
dgeIdc <- DGEList(counts = as.matrix(expIdc), samples = phenoIdc, group = phenoIdc$Group)
```
```{r}
dim(dgeIlc)
dim(dgeIdc)
```
# Prepares and assigns gene annotations to the DGEList
```{r}
geneid <- rownames(dgeIdc)
genes <- ensembldb::select(EnsDb.Hsapiens.v86, 
                           keys=geneid, 
                           columns=c("ENTREZID", "SYMBOL", "GENEID",
                                     "TXSEQSTART", "TXSEQEND", "SEQSTRAND"),
                           keytype="GENEID")
genes <- genes[!duplicated(genes$GENEID),]
dgeIlc$genes <- genes
dgeIdc$genes <- genes
```
```{r}
head(genes)
```
# Filtering lowly expressed genes
```{r}
keepIlc <- filterByExpr(dgeIlc, group = dgeIlc$samples$Group)
dgeIlc <- dgeIlc[keepIlc, , keep.lib.sizes = FALSE]

keepIdc <- filterByExpr(dgeIdc, group = dgeIdc$samples$Group)
dgeIdc <- dgeIdc[keepIdc, , keep.lib.sizes = FALSE]
```
```{r}
dim(dgeIlc)
dim(dgeIdc)
```
# Normalizing
```{r}
dgeIlc <- calcNormFactors(dgeIlc, method = "TMM")
dgeIdc <- calcNormFactors(dgeIdc, method = "TMM")
```
# MDS plot of ILC
```{r}
lcpmIlc <- cpm(dgeIlc, log = TRUE)
colGroup <- as.factor(dgeIlc$samples$Group)
levels(colGroup) <-  brewer.pal(nlevels(colGroup), "Set1")
colGroup <- as.character(colGroup)
tiff("./Figures/mds-normal-ilc-unpaired.tiff", height = 4, width = 4,
     units = "in", res = 300)
par(mar = c(5, 4, 2, 2) + 0.1)
plotMDS(lcpmIlc, labels = dgeIlc$samples$Group, col = colGroup,
        xlab = "logFC Dimension 1", ylab = "logFC Dimension 2")
title(main = "MDS Plot of Unpaired Normal and ILC")
dev.off()
```
# MDS plot of IDC
```{r}
lcpmIdc <- cpm(dgeIdc, log = TRUE)
colGroup <- as.factor(dgeIdc$samples$Group)
levels(colGroup) <-  brewer.pal(nlevels(colGroup), "Set1")
colGroup <- as.character(colGroup)
tiff("./Figures/mds-normal-idc-unpaired.tiff", height = 4, width = 4,
     units = "in", res = 300)
par(mar = c(5, 4, 2, 2) + 0.1)
plotMDS(lcpmIdc, labels = dgeIdc$samples$Group, col = colGroup,
        xlab = "logFC Dimension 1", ylab = "logFC Dimension 2")
title(main = "MDS Plot of Unpaired Normal and IDC")
dev.off()
```
# Design matrix of normal vs. ILC
```{r}
designIlc <- model.matrix(~0 + Group, data = dgeIlc$samples)
colnames(designIlc) <- str_remove_all(colnames(designIlc), "Group")
contrIlc <- makeContrasts(TumorVsNormal = TP-NT, levels = designIlc)
contrIlc
```
# Voom
```{r}
par(mfrow = c(1,1))
v <- voom(dgeIlc, designIlc, plot=TRUE)
```
# DGE Analysis normal vs.ILC
```{r}
vfit <- lmFit(v, designIlc)
vfit <- contrasts.fit(vfit, contrasts = contrIlc)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```
# Filtering top DEGs in normal vs. ILC
```{r}
topIlc <- topTreat(efit, coef = 1, n = Inf)
topSigIlc <- topIlc[topIlc$adj.P.Val < 0.05 & abs(topIlc$logFC) > 2, ]
topSigIlc <- topSigIlc[order(topSigIlc$adj.P.Val), ]
head(topSigIlc)
```
```{r}
dim(topSigIlc)
```
# Volcano plot of DEGs in normal vs. ILC
```{r}
EnhancedVolcano(topIlc, x = "logFC", y = "adj.P.Val", lab = topIlc$SYMBOL,
                FCcutoff = 2, pCutoff = 0.05, cutoffLineType = 'twodash',
                cutoffLineWidth = 0.8, pointSize = 1.0, legendPosition = 'none',
                title = "", subtitle = "", caption = "", labSize = 3) +
  theme_pubr(legend = "none")
ggsave("./Figures/volcano-normal-ilc-unpaired.tiff", height = 6, width = 4, dpi = 300)
```
# Saving DEGs in normal vs. ILC
```{r}
write.csv(topIlc, "./Results/dge-normal-vs-ilc-unpaired-all.csv", row.names = FALSE)
write.csv(topSigIlc, "./Results/dge-normal-vs-ilc-unpaired-sig.csv", row.names = FALSE)
```
# Design matrix of normal vs. IDC
```{r}
designIdc <- model.matrix(~0 + Group, data = dgeIdc$samples)
colnames(designIdc) <- str_remove_all(colnames(designIdc), "Group")
contrIdc <- makeContrasts(TumorVsNormal = TP-NT, levels = designIdc)
contrIdc
```
# Voom
```{r}
par(mfrow = c(1, 1))
v <- voom(dgeIdc, designIdc, plot=TRUE)
```
# DGE analysis of normal vs. IDC
```{r}
vfit <- lmFit(v, designIdc)
vfit <- contrasts.fit(vfit, contrasts = contrIdc)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```
# Filtering top DEGs in normal vs. IDC
```{r}
topIdc <- topTreat(efit, coef = 1, n = Inf)
topSigIdc <- topIdc[topIdc$adj.P.Val < 0.05 & abs(topIdc$logFC) > 2, ]
topSigIdc <- topSigIdc[order(topSigIdc$adj.P.Val), ]
head(topSigIdc)
```
```{r}
dim(topSigIdc %>% filter(logFC < 0))
dim(topSigIdc %>% filter(logFC > 0))
```
# Volcano plot of DEGs in normal vs. IDC
```{r}
EnhancedVolcano(topIdc, x = "logFC", y = "adj.P.Val", lab = topIdc$SYMBOL,
                FCcutoff = 2, pCutoff = 0.05, cutoffLineType = 'twodash',
                cutoffLineWidth = 0.8, pointSize = 1.0, legendPosition = 'none',
                title = "", subtitle = "", caption = "", labSize = 3) +
  theme_pubr(legend = "none")
ggsave("./Figures/volcano-normal-idc-unpaired.tiff", height = 6, width = 4, dpi = 300)
```
# Saving DEGs in normal vs. IDC
```{r}
write.csv(topIdc, "./Results/dge-normal-vs-idc-unpaired-all.csv", row.names = FALSE)
write.csv(topSigIdc, "./Results/dge-normal-vs-idc-unpaired-sig.csv", row.names = FALSE)
```
# Clears workspace
```{r}
rm(list = ls())
```
